Categories of Optics

### 2018-07-10 paper.tex
More quibbles

\documentclass[11pt,letterpaper]{article}
\usepackage{authblk}
% \usepackage[utf8]{inputenc}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\let\amssquare\square
\usepackage{mathtools}
% \usepackage{stmaryrd}
% \usepackage{tensor}
\usepackage[mathscr]{eucal}
\usepackage{url}
% \usepackage{marvosym}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
%\usepackage{geometry}
\usepackage{minted}
\usepackage{cite}

@@

For a pair of maps $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\rep{l}{r} : (S, S') \hto (A, A')$ for their image in $\Optic_\otimes((S, S'), (A, A'))$, and say that $M$ is the residual for this representative. Optics will always be written with a crossed arrow $\hto$, to distinguish them from morphisms of $\C$.

@@

\begin{example}
  For any three objects $M, A, A' \in \C$, there is the \emph{tautological} optic \[t_{M,A} : (M \otimes A, M \otimes A') \hto (A, A')\] which is given by $\rep{\id_{M \otimes A}}{\id_{M \otimes A}}$.
\end{example}

@@

If $\C$ is a category with finite products, then $\Lens = \Optic_\times$ is exactly the category of lenses described in the introduction. That optics correspond to pairs of $\fget$ and $\fput$ functions is shown via the following isomorphisms:
\begin{align*}
  \Optic_\times((S, S'), (A, A')) 
  &= \int^{M \in \C} \C(S, M \times A) \times \C(M \times A', S') \\
  &\cong \int^{M \in \C} \C(S, M) \times \C(S, A) \times \C(M \times A', S') && \text{(universal property of product)} \\
  &\cong \C(S, A) \times \C(S \times A', S') && \text{(Yoneda reduction)}
\end{align*}
Let $\conc_\times((S, S'), (A, A'))$ denote this set of concrete lenses. Explicitly the isomorphism states that, given an optic $\rep{l}{r} : (S, S') \hto (A, A')$, the corresponding concrete lens is the pair $(\fget, \fput)$, where $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $(\fget, \fput)$, the corresponding optic has representative $\rep{[\id_S, \fget]}{\fput}$.

Using the isomorphism in both directions, we leave it to the reader to verify that composition in $\Optic_\times$ corresponds to ordinary composition of concrete lenses. (Of course, there is only one possible way such a collection of concrete maps could be composed!)

@@

\begin{remark}
This definition of teleological category differs from the original one, in that the duality switches the order of the tensor product. We do this so that compact closed category are teleological, but the bookkeeping becomes more confusing.
\end{remark}

@@

\begin{definition}
  A \emph{teleological functor} $F : \C \to \D$ is a symmetric monoidal functor that restricts to a functor $F_d : \C_d \to \D_d$ on the dualisable subcategories, commutes with the duality and such that the counits are preserved:
  \[
   \begin{tikzcd}
    F(X \otimes X^*) \ar[r, "\phi_{X, X^*}"]  \ar[d, "F\varepsilon_X", swap] & FX \otimes FX^* \ar[d, "\varepsilon_{FX}"] \\
    FI \ar[r, "\phi", swap] & I
  \end{tikzcd}
  \]
\end{definition}

%\begin{definition}
%  A \emph{teleological natural transformation} $\alpha : F \Rightarrow G$ is a monoidal natural transformation that is additionally compatible with the dualisation:
%  \[
%  \begin{tikzcd}
%    (FX)^* \ar[r, "\cong"]  \ar[d, "(\alpha_X)^*", swap] & F(X^*) \ar[d, "\alpha_{X^*}"] \\
%    (GX)^* \ar[r, "\cong", swap] & G(X^*)
%  \end{tikzcd}
%  \]
%  whenever $\alpha_X$ is dualisable. \todo{Does adding this condition make sense?}
%\end{definition}

Together we have $\Tele$, the subcategory of $\SymmMonCat$ given by teleological categories and teleological functors. There are evident functors $U : \Tele \to \SymmMonCat$ and ${(-)}_d : \Tele \to \SymmMonCat$, that take a teleological category to its underlying symmetric monoidal category and subcategory of dualisable morphisms respectively.

@@

\begin{theorem}\label{optic-is-free-teleological-cat}
  Suppose $\C$ is a strict symmetric monoidal category and $\T$ is a teleological category. For all symmetric monoidal functors $F : \C \to \T_d$, there exists a teleological functor $K : \Optic_\otimes \to \T$, unique up to teleological natural isomorphism with the property $Kj \cong F$.
\end{theorem}
\begin{proof}
  We begin by constructing $K$. Note that any $(S, S')$ in $\Optic_\otimes$ can be written as $j(S) \switched {j(S')}^*$, so on objects define $K(S, S') = FS \otimes {(FS')}^*$. Suppose $\rep{l}{r} : (S, S') \hto (A, A')$ is an optic. By the previous Proposition,
  \begin{align*}
    \rep{l}{r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched {j(rs_{A', M})}^*)
  \end{align*}
  So up to isomorphism we are forced to define
  \begin{align*}
    K\rep{l}{r} = (FA \otimes \varepsilon_{FM} \otimes {(FA')}^*)(F(s_{M,A}l) \otimes {(F(rs_{A', M}))}^* )
  \end{align*}
  The diagram for $K\rep{l}{r}$ in $\T$ is as follows:
  \begin{center}
    \input{diagrams/k-generic-optic.tikz}
  \end{center}

@@

There are now a lot of things to check.
\begin{itemize}
\item Well-definedness: Suppose we have two optics related by the coend relation:
  \begin{align*}
    \rep{(f \otimes A) l}{r} = \rep{l}{r (f \otimes A')}
  \end{align*}
  Then well-definedness is shown by the equivalence of diagrams
  \begin{center}
    \input{diagrams/k-well-defined-left.tikz}
    \qquad \raisebox{1.5cm}{$=$} \qquad
    \input{diagrams/k-well-defined-right.tikz}
  \end{center}
  using naturality of the symmetry and extranaturality of the counit.
\item Functoriality: We have an equivalence of diagrams
  \begin{center}
    \input{diagrams/k-functorial-left.tikz}
    \quad \raisebox{1.5cm}{$=$} \quad
    \input{diagrams/k-functorial-right.tikz}
  \end{center}
using naturality of the symmetry and monoidality of the counit.
\item Preservation of duals:
\begin{align*}
K({(S, S')}^*)
= K(S', S)
= FS' \otimes {(FS)}^*
\cong {(FS \otimes {(FS')}^*)}^*
= {(K(S, S'))}^*
\end{align*}
\item Monoidality: We have the isomorphism \[ \phi_{(S, S'),(T, T')} : K((S, S') \switched (T, T')) \to K(S, S') \otimes K(T, T')\] given by the composite
\begin{align*}
K((S, S') \switched (T, T'))
&= K(S \otimes T, T' \otimes S) \\
&= F(S \otimes T) \otimes {F(T' \otimes S')}^* \\
&\xrightarrow{\phi_{S, T} \otimes \phi_{T', S'}^*} FS \otimes FT \otimes {(FS')}^* \otimes {(FT')}^* \\
&\xrightarrow{FS \otimes s \otimes (FT')^*} FS \otimes {(FS')}^* \otimes FT \otimes {(FT')}^* \\
&= K(S, S') \otimes K(T, T')
\end{align*}
using the symmetry of $\T$, and
\begin{align*}
K(I, I)
&= FI \otimes {(FI)}^* \\
&\cong I \otimes I^* \\
&\cong I
\end{align*}
That these obey the required coherences is straightforward. More difficult is showing that the first isomorphism is natural in $(S, S')$ and $(T, T')$: \todo{todo}
\item Preservation of dualisable morphisms:  For a morphism $\iota(f, g)$:
\begin{align*}
K(\iota(f, g))
    &= K(\rep{\lambda_A^{-1} f}{g \lambda_{A'}}) \\
    &= (FA \otimes \varepsilon_{FI} \otimes {(FA')}^*)(F(s_{I,A}\lambda_A^{-1} f) \otimes (F(g \lambda_{A'}s_{A', I}))^* ) \\
    &= (FA \otimes {(FA')}^*)(Ff \otimes {(Fg)}^* ) \\
    &= Ff \otimes {(Fg)}^*
\end{align*}
and this is dualisable, as dualisability is preserved by taking the monoidal product and duals.
\item Preservation of counits:
\begin{align*}
K(\varepsilon_{(S, S')})
&= K(c_{S \otimes S'}) \\
&= K(\rep{\rho_{S \otimes S'}^{-1}}{\rho_{S \otimes S'}}) \\
&= (FI \otimes \varepsilon_{F(S \otimes S')} \otimes (FI)^*)(F(s_{S \otimes S',I}\rho_{S \otimes S'}^{-1}) \otimes (F(\rho_{S \otimes S'} s_{I, S \otimes S'}))^* ) \\
&= (\varepsilon_{F(S \otimes S')})(F(S \otimes S') \otimes F(S \otimes S')^* ) \\
&= \varepsilon_{F(S \otimes S')}
\end{align*}
by strictness of the monoidal unit. This morphism is indeed equal to the composite $\varepsilon_{K(S, S')} \phi_{(S, S'), {(S, S')}^*}$, using extranaturality to commute the structure maps $\phi_{S, S'}$ of $F$ and the symmetry of $\T$ through $\varepsilon$. \todo{write details?}
\end{itemize}
\end{proof}

\begin{theorem}
$\Optic : \SymmMonCat \to \Tele$ is left biadjoint to the underlying dualisable morphisms functor ${(-)}_d : \Tele \to \SymmMonCat$
\end{theorem}
\begin{proof}
\todo{figure out exactly what else needs proving}
\end{proof}

@@

Without loss of generality, we just check the case of $\repthree{l;\phi_S}{c}{r} = \repthree{l}{\phi_S ; c}{r}$:
\begin{align*}
&\repthree{(l;\phi_S \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l');(\phi_S \otimes N \otimes B);(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l');(\phi_S \otimes N \otimes B)}{(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l')}{(\phi_S \otimes N \otimes B);(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l')}{(\phi_S;c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l');(M' \otimes s_{A,N} \otimes B)}{(M' \otimes s_{A,N} \otimes B);(\phi_S;c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots}
\end{align*}
So again, we can transplant the chain of relations proving $\repthree{l}{r;l}{r} = \repthree{l}{\id_{MA}}{r}$ and $\repthree{l'}{r';l'}{r'} = \repthree{l'}{\id_{NB}}{r'}$ to prove:
\begin{align*}
&\twice(\rep{l}{r} \otimes \rep{l'}{r'}) \\
&= \twice(\rep{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(r \otimes r')}) \\
&=\repthree{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(r \otimes r');(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(r \otimes r')} \\
&= \repthree{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(rl \otimes r'l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(r \otimes r')} \\
&= \repthree{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(\id_{M \otimes A} \otimes \id_{N \otimes B});(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(r \otimes r')} \\
&= \repthree{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B);(r \otimes r')} \\
&= \repthree{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{\id_{M \otimes N \otimes A \otimes B}}{(M \otimes s_{A,N} \otimes B);(r \otimes r')} \\
&= \once(\rep{l}{r} \otimes \rep{l'}{r'})
\end{align*}

@@

\begin{proposition}
If $T : \C \to \C$ is a strong monad then $\C$ acts on $\C_T$ by $X \cdot Y \mapsto X \otimes Y$.
\end{proposition}

@@

A morphism $f : A \to B$ in $\C$ induces a natural transformation $A \otimes - \Rightarrow B \otimes -$ with components $A \otimes X \to T(B \otimes X)$ given by composing $A \otimes X \to B \otimes X$ with the unit of the monad. Naturality follows by the naturality of the strength and the unit of $T$.
%Naturality is easy to check:
%\[\begin{tikzcd}
%A \otimes X \ar[r] \ar[d] & B \otimes X \ar[r] \ar[d] & T(B \otimes X) \ar[d] \\
%A \otimes TY \ar[r] \ar[d] & B \otimes TY \ar[r] \ar[d] & T(B \otimes TY) \ar[d] \\
%T(A \otimes Y) \ar[r] & T(B \otimes Y) \ar[r] & TT(B \otimes Y)
%\end{tikzcd}\]
%The upper left square commutes by functoriality of $\otimes$, lower left by naturality of the strength, the two right squares by naturality of the unit.

@@

\begin{proposition}
A concrete effectful lens is lawful iff
  \begin{align*}
    \mu_S T(\ev_{A, TS}) \; \munzip &= \eta_S \\
    T(\eta_{\homC(A, TS) \otimes A}\coev_{\homC(A, TS), A} \otimes A)\munzip &= T((\munzip \circ_T -) \otimes A)\munzip
  \end{align*}
  where \[ \munzip \circ_T - : \homC(A, TS) \to \homC(A, T(\homC(A, TS) \otimes A)) \] denotes internal Kleisli composition and \[\coev_{\homC(A, TS), A} : \homC(A, TS) \to \homC(A, \homC(A, TS) \otimes A) \] is coevaluation.
\end{proposition}
  Or, if you prefer do-notation:
\begin{minted}{haskell}
do (c, a) <- munzip s
   c a
==
return s

do (c, a) <- munzip s
   let f a' = do
     s' <- c a'
     munzip s'
   return (f, a)
==
do (c, a) <- munzip s
   let f a' = (c, a')
   return (f, a)
\end{minted}
\begin{proof}
This is exactly analogous to Proposition~\ref{prop:concrete-linear-lawful} for linear lenses, with the unit and multiplication of the monad inserted in the appropriate places.
%\begin{align*}
%&\int^{M, N \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, N \monact A) \times \C_T(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, N \monact A) \times \C(N, \homC(A, TS)) \\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, \homC(A, TS) \monact A)\\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M \times A, T(\homC(A, TS) \times A))\\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M, \homC(A, T(\homC(A, TS) \monact A)))\\
%&\cong \C_T(S, \homC(A, T(\homC(A, TS) \times A)) \times A) \\
%&\cong \C(S, T(\homC(A, T(\homC(A, TS) \times A)) \times A))
%\end{align*}
\end{proof}

@@

%\subsubsection{Partial Lenses}
%
%\begin{definition}
%The \emph{partiality} or \emph{maybe monad} is defined by
%\begin{align*}
%P X = X \sqcup 1
%\end{align*}
%\todo{todo: define its unit, multiplication and strength}
%\end{definition}

@@

\subsection{Learners}

>>>>

\documentclass[11pt,letterpaper]{article}
\usepackage{authblk}
% \usepackage[utf8]{inputenc}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\let\amssquare\square
\usepackage{mathtools}
% \usepackage{stmaryrd}
% \usepackage{tensor}
\usepackage[mathscr]{eucal}
\usepackage{url}
% \usepackage{marvosym}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
%\usepackage{geometry}
\usepackage{minted}
\usepackage{cite}
\usepackage{multicol}

@@

For a pair of maps $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\rep{l}{r} : (S, S') \hto (A, A')$ for their image in $\Optic_\otimes((S, S'), (A, A'))$, and say that $M$ is the residual for this representative. Optics will always be written with a crossed arrow $\hto$ to distinguish them from morphisms of $\C$.

\begin{example}
  For any three objects $M, A, A' \in \C$, there is the \emph{tautological} optic \[t_{M,A} : (M \otimes A, M \otimes A') \hto (A, A')\] given by $\rep{\id_{M \otimes A}}{\id_{M \otimes A'}}$.
\end{example}

@@

@@

\begin{remark}
If $\C$ is a category with finite products, then $\Lens = \Optic_\times$ is exactly the category of lenses described in the introduction, so long as we restrict to optics of the form $(S, S) \hto (A, A)$. We see that optics correspond to pairs of $\fget$ and $\fput$ functions via the following isomorphisms:
\begin{align*}
  \Optic_\times((S, S'), (A, A')) 
  &= \int^{M \in \C} \C(S, M \times A) \times \C(M \times A', S') \\
  &\cong \int^{M \in \C} \C(S, M) \times \C(S, A) \times \C(M \times A', S') && \text{(universal property of product)} \\
  &\cong \C(S, A) \times \C(S \times A', S') && \text{(Yoneda reduction)}
\end{align*}
Let $\conc_\times((S, S'), (A, A'))$ denote this set of concrete lenses. Explicitly the isomorphism states that, given an optic $\rep{l}{r} : (S, S') \hto (A, A')$, the corresponding concrete lens is the pair $(\fget, \fput)$, where $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $(\fget, \fput)$, the corresponding optic has representative $\rep{[\id_S, \fget]}{\fput}$.

Using the isomorphism in both directions, we leave it to the reader to verify that composition in $\Optic_\times$ corresponds to ordinary composition of concrete lenses. (Of course, there is only one possible way such a collection of concrete maps could be composed!)
\end{remark}

@@

\begin{remark}
This definition of teleological category differs from the original one, in that the duality switches the order of the tensor product. We do this so that compact closed categories are teleological, but the bookkeeping does admittedly become more confusing.
\end{remark}

@@

\begin{definition}
  A \emph{teleological functor} $F : \C \to \D$ is a symmetric monoidal functor that restricts to a functor $F_d : \C_d \to \D_d$ on the dualisable subcategories, commutes with the duality via a monoidal natural isomorphism $d_X : F(X^*) \to (FX)^*$, and such that the counits are preserved:
  \[
   \begin{tikzcd}[column sep = large]
    F(X \otimes X^*) \ar[r, "\phi_{X, X^*}"]  \ar[d, "F\varepsilon_X", swap] & FX \otimes F(X^*) \ar[r, "FX \otimes d_X"] & FX \otimes (FX)^* \ar[d, "\varepsilon_{FX}"] \\
    FI \ar[rr, "\phi", swap] & & I
  \end{tikzcd}
  \]

  A \emph{teleological natural transformation} $\alpha : F \Rightarrow G$ is a monoidal natural transformation whose components are all dualisable and that is additionally compatible with the dualisation:
  \[
  \begin{tikzcd}
    (FX)^* \ar[r, "\cong"]  \ar[d, "(\alpha_X)^*", swap] & F(X^*) \ar[d, "\alpha_{X^*}"] \\
    (GX)^* \ar[r, "\cong", swap] & G(X^*)
  \end{tikzcd}
  \]
\end{definition}

Together we have $\Tele$, the sub-2-category of $\SymmMonCat$ given by teleological categories, teleological functors and teleological natural transformations. There are evident strict 2-functors $U : \Tele \to \SymmMonCat$ and ${(-)}_d : \Tele \to \SymmMonCat$, that take a teleological category to its underlying symmetric monoidal category and subcategory of dualisable morphisms respectively.

@@

\begin{theorem}\label{optic-is-free-teleological-cat}
  Suppose $\C$ is a strict symmetric monoidal category and $\T$ is a teleological category. For all symmetric monoidal functors $F : \C \to \T_d$, there exists a teleological functor $K : \Optic_\otimes \to \T$, unique up to teleological natural isomorphism with the property $Kj \cong F$.
  Furthermore, $\Optic : \SymmMonCat \to \Tele$ is left biadjoint to the underlying dualisable morphisms functor ${(-)}_d : \Tele \to \SymmMonCat$.
\end{theorem}
\begin{proof}
  \todo{Notes: I think uniqueness is easy but I don't know how best to phrase it. I might instead be able to do something 1-categorical, so having the property $Kj = F$ strictly. This would possibly be easier but would involve using strict monoidal functors everywhere that strictly preserve duals, and having $I^* = I$. This is true for $\Optic$, but is that actually what we want in general? The teleological category definition is so contrived I can probably just make it whatever I need it to be, but that's a bit unsatisfying.}

  We begin by constructing $K$. Note that any $(S, S')$ in $\Optic_\otimes$ can be written as $j(S) \switched {j(S')}^*$, so we define $K(S, S') = FS \otimes {(FS')}^*$. Suppose $\rep{l}{r} : (S, S') \hto (A, A')$ is an optic. By the previous Proposition,
  \begin{align*}
    \rep{l}{r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched {j(rs_{A', M})}^*)
  \end{align*}
  So in order for $Kj \cong F$ to hold, we define
  \begin{align*}
    K\rep{l}{r} = (FA \otimes \varepsilon_{FM} \otimes {(FA')}^*)(F(s_{M,A}l) \otimes {(F(rs_{A', M}))}^* )
  \end{align*}
  The diagram for $K\rep{l}{r}$ in $\T$ is as follows:
  \begin{center}
    \input{diagrams/k-generic-optic.tikz}
  \end{center}

@@

There are now a lot of things to check.
\begin{itemize}
\item Well-definedness: Suppose we have two optics related by the coend relation:
  \begin{align*}
    \rep{(f \otimes A) l}{r} = \rep{l}{r (f \otimes A')}
  \end{align*}
  Then well-definedness is shown by the equivalence of diagrams
  \begin{center}
    \input{diagrams/k-well-defined-left.tikz}
    \qquad \raisebox{1.5cm}{$=$} \qquad
    \input{diagrams/k-well-defined-right.tikz}
  \end{center}
  using naturality of the symmetry and extranaturality of the counit.
\item Functoriality: We have an equivalence of diagrams
  \begin{center}
    \input{diagrams/k-functorial-left.tikz}
    \quad \raisebox{1.5cm}{$=$} \quad
    \input{diagrams/k-functorial-right.tikz}
  \end{center}
using naturality of the symmetry and monoidality of the counit.
\item Preservation of duals:
\begin{align*}
K({(S, S')}^*)
= K(S', S)
= FS' \otimes {(FS)}^*
\cong {(FS \otimes {(FS')}^*)}^*
= {(K(S, S'))}^*
\end{align*}
\item Monoidality: We have the isomorphism \[ \phi_{(S, S'),(T, T')} : K((S, S') \switched (T, T')) \to K(S, S') \otimes K(T, T')\] given by the composite
\begin{align*}
K((S, S') \switched (T, T'))
&= K(S \otimes T, T' \otimes S) \\
&= F(S \otimes T) \otimes {F(T' \otimes S')}^* \\
&\xrightarrow{\phi_{S, T} \otimes \phi_{T', S'}^*} FS \otimes FT \otimes {(FS')}^* \otimes {(FT')}^* \\
&\xrightarrow{FS \otimes s \otimes (FT')^*} FS \otimes {(FS')}^* \otimes FT \otimes {(FT')}^* \\
&= K(S, S') \otimes K(T, T')
\end{align*}
using the symmetry of $\T$, and
\begin{align*}
K(I, I)
&= FI \otimes {(FI)}^* \\
&\cong I \otimes I^* \\
&\cong I
\end{align*}
That these obey the required coherences is straightforward. More difficult is showing that the first isomorphism is natural in $(S, S')$ and $(T, T')$: \todo{todo}
\item Preservation of dualisable morphisms:  For a morphism $\iota(f, g)$:
\begin{align*}
K(\iota(f, g))
    &= K(\rep{\lambda_A^{-1} f}{g \lambda_{A'}}) \\
    &= (FA \otimes \varepsilon_{FI} \otimes {(FA')}^*)(F(s_{I,A}\lambda_A^{-1} f) \otimes (F(g \lambda_{A'}s_{A', I}))^* ) \\
    &= (FA \otimes {(FA')}^*)(Ff \otimes {(Fg)}^* ) \\
    &= Ff \otimes {(Fg)}^*
\end{align*}
and this is dualisable, as dualisability is preserved by taking the monoidal product and duals.
\item Preservation of counits:
\begin{align*}
K(\varepsilon_{(S, S')})
&= K(c_{S \otimes S'}) \\
&= K(\rep{\rho_{S \otimes S'}^{-1}}{\rho_{S \otimes S'}}) \\
&= (FI \otimes \varepsilon_{F(S \otimes S')} \otimes (FI)^*)(F(s_{S \otimes S',I}\rho_{S \otimes S'}^{-1}) \otimes (F(\rho_{S \otimes S'} s_{I, S \otimes S'}))^* ) \\
&= (\varepsilon_{F(S \otimes S')})(F(S \otimes S') \otimes F(S \otimes S')^* ) \\
&= \varepsilon_{F(S \otimes S')}
\end{align*}
by strictness of the monoidal unit. This morphism is indeed equal to the composite $\varepsilon_{K(S, S')} \phi_{(S, S'), {(S, S')}^*}$, using extranaturality to commute the structure maps $\phi_{S, S'}$ of $F$ and the symmetry of $\T$ through $\varepsilon$. \todo{write details?}
\end{itemize}

To conclude biadjointness it remains to show that for any symmetric monoidal category $\C$ and teleological category $\T$, restriction along $j : \C \to \Optic_\C$ defines an equivalence of categories \[\Tele(\Optic_\C, \T) \simeq \SymmMonCat(\C, \T_d). \]
The restriction of a functor along $j$ indeed has its image in the subcategory $\T_d$, as by definition $j(f) = \iota(f, I)$ is dualisable. And given two teleological functors $K, L : \Optic_\C \to \T$, the whiskering of a teleological natural transformation $\alpha : K \Rightarrow L$ by $j$ is a well defined natural transformation $\beta : K j \Rightarrow L j$ in $\SymmMonCat(\C, \T_d)$ as the components of teleological natural transformations must be dualisable.

We provide an inverse to this whiskering operation as follows. If $\beta : K j \Rightarrow L j$ is a monoidal natural transformation, define $\alpha : K \Rightarrow L$ to be the natural transformation with components:
\begin{align*}
K(S, S') = K(j(S) \switched {j(S')}^*) \xrightarrow{\phi} Kj(S) \otimes K({j(S')}^*) \xrightarrow{\beta_S \otimes \beta_{S'}^*} Lj(S) \otimes L({j(S')}^*) \xrightarrow{\phi^{-1}} L(j(S) \switched {j(S')}^*) = L(S, S')
\end{align*}
\todo{Not quite right, we should be using $d_X$} We leave showing that this is an inverse to the reader.
\end{proof}

\begin{remark}
The above theorem and its proof have much in common with \cite[Proposition 5.2]{JoyalStreetVerity}, which gave a similar universal property for their $\mathrm{Int}$ construction on traced monoidal categories.
\end{remark}

@@

Without loss of generality, we just check the case of $\repthree{l;\phi_S}{c}{r} = \repthree{l}{\phi_S ; c}{r}$. (The right entry is identical in all expressions, so we omit it.)
\begin{align*}
&\repthree{(l;\phi_S \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l');(\phi_S \otimes N \otimes B);(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l');(\phi_S \otimes N \otimes B)}{(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l')}{(\phi_S \otimes N \otimes B);(c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l')}{(\phi_S;c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{(l \otimes l');(M' \otimes s_{A,N} \otimes B)}{(M' \otimes s_{A,N} \otimes B);(\phi_S;c \otimes c');(M \otimes s_{A,N} \otimes B)}{\dots}
\end{align*}
So again, we can transplant the chain of relations proving $\repthree{l}{r;l}{r} = \repthree{l}{\id_{MA}}{r}$ and $\repthree{l'}{r';l'}{r'} = \repthree{l'}{\id_{NB}}{r'}$ to prove:
\begin{align*}
&\twice(\rep{l}{r} \otimes \rep{l'}{r'}) \\
&= \twice(\rep{(l \otimes l');(M \otimes s_{A,N} \otimes B)}{(M \otimes s_{A,N} \otimes B');(r \otimes r')}) \\
&=\repthree{\dots}{(M \otimes s_{A,N} \otimes B);(r \otimes r');(l \otimes l');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{\dots}{(M \otimes s_{A,N} \otimes B);(rl \otimes r'l');(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{\dots}{(M \otimes s_{A,N} \otimes B);(\id_{M \otimes A} \otimes \id_{N \otimes B});(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{\dots}{(M \otimes s_{A,N} \otimes B);(M \otimes s_{A,N} \otimes B)}{\dots} \\
&= \repthree{\dots}{\id_{M \otimes N \otimes A \otimes B}}{\dots} \\
&= \once(\rep{l}{r} \otimes \rep{l'}{r'})
\end{align*}

@@

\begin{proposition}
If $T : \C \to \C$ is a strong monad then $\C$ acts on $\C_T$ by $X \cdot Y := X \otimes Y$.
\end{proposition}

@@

A morphism $f : A \to B$ in $\C$ induces a natural transformation $A \otimes - \Rightarrow B \otimes -$ of functors $\C_T \to \C_T$, with components $A \otimes X \to T(B \otimes X)$ given by composing $A \otimes X \to B \otimes X$ with the unit of the monad. Naturality follows by the naturality of the strength and the unit of $T$.
%Naturality is easy to check:
%\[\begin{tikzcd}
%A \otimes X \ar[r] \ar[d] & B \otimes X \ar[r] \ar[d] & T(B \otimes X) \ar[d] \\
%A \otimes TY \ar[r] \ar[d] & B \otimes TY \ar[r] \ar[d] & T(B \otimes TY) \ar[d] \\
%T(A \otimes Y) \ar[r] & T(B \otimes Y) \ar[r] & TT(B \otimes Y)
%\end{tikzcd}\]
%The upper left square commutes by functoriality of $\otimes$, lower left by naturality of the strength, the two right squares by naturality of the unit.

@@

\begin{proposition}
A concrete effectful lens is lawful iff
  \begin{align*}
    \mu_S T(\ev_{A, TS}) \; \munzip &= \eta_S \\
    T(\eta_{\homC(A, TS) \otimes A}\coev_{\homC(A, TS), A} \otimes A)\munzip &= T((\munzip \circ_T -) \otimes A)\munzip
  \end{align*}
  where \[ \munzip \circ_T - : \homC(A, TS) \to \homC(A, T(\homC(A, TS) \otimes A)) \] denotes internal Kleisli composition and \[\coev_{\homC(A, TS), A} : \homC(A, TS) \to \homC(A, \homC(A, TS) \otimes A) \] is coevaluation.
\end{proposition}
  Or, if you prefer do-notation, the two laws are:
\begin{multicols}{2}
\begin{minted}{haskell}
do (c, a) <- munzip s
   c a
==
return s
\end{minted} 
~\columnbreak
\begin{minted}{haskell}
do (c, a) <- munzip s
   let f a' = do
     s' <- c a'
     munzip s'
   return (f, a)
==
do (c, a) <- munzip s
   let f a' = (c, a')
   return (f, a)
\end{minted}
\end{multicols}

\begin{proof}
This is exactly analogous to Proposition~\ref{prop:concrete-linear-lawful} for linear lenses, with the unit and multiplication of the monad inserted in the appropriate places.
%\begin{align*}
%&\int^{M, N \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, N \monact A) \times \C_T(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, N \monact A) \times \C(N, \homC(A, TS)) \\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, \homC(A, TS) \monact A)\\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M \times A, T(\homC(A, TS) \times A))\\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M, \homC(A, T(\homC(A, TS) \monact A)))\\
%&\cong \C_T(S, \homC(A, T(\homC(A, TS) \times A)) \times A) \\
%&\cong \C(S, T(\homC(A, T(\homC(A, TS) \times A)) \times A))
%\end{align*}
\end{proof}

@@

%\subsubsection{Partial Lenses}
%
%The simplest nontrivial monad we could try is the maybe monad.
%
%\begin{definition}
%The \emph{partiality} or \emph{maybe monad} is defined by
%\begin{align*}
%T X = X \sqcup 1
%\end{align*}
%with unit $\eta_X : X \to X \sqcup 1$ the inclusion, multiplication $\mu_X : (X \sqcup 1) \sqcup 1 \to X \sqcup 1$ given by the fold map $1 \sqcup 1 \to 1$, and strength $\theta_{A, B}$ given by the composite
%\[
%A \times (B \sqcup 1) \to (A \times B) \sqcup (A \times 1) \to (A \times B) \sqcup 1
%\]
%using the unique map $A \times 1 \to 1$.
%\end{definition}
%

@@

\subsection{Learners}
A recent paper \cite{BackpropAsFunctor} in applied category theory describes a compositional approach to machine learning, with a category whose morphisms describe learning algorithms.
